---
title: "Bertke_Reactivation"
author: "Seth Frietze"
date: "6/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HSV1 reactivation in primary mouse neurons  

Herpes simplex viruses 1 and 2 (HSV-1 and HSV-2) establish latency in sensory and autonomic neurons, from which they can reactivate to cause recurrent disease such as cold, genital, and ocular sores and encephalitis.  However, some types of mature neurons support HSV replication upon entry, while other types of neurons inhibit viral replication, resulting in latency. Exogenous stimuli can trigger reactivation but only from a portion of latently infected neurons. The neuronal populations that support these divergent outcomes differ for HSV-1/HSV-2 are not known, leading to different anatomical patterns and frequencies of recurrent disease. To better understand factors that differentiate neurons in their ability to permit or prevent reactivation in response to exogenous stimuli, we performed RNA sequencing on reactivating and non-reactivating populations of adult murine neurons following neurotrophic factor deprivation.


SMARTer® cDNA Library Construction, Illumina, STAR mm10, Gencode mV14, counts table
```{r load data, message=FALSE}
library(tidyverse)
library(DESeq2)
#read HTseq files
countdata <- read.table("~/Documents/data/Bertke/2020/gene_read_counts_table_all_final2.tsv", header=TRUE, row.names=1)

colnames(countdata) <- colnames(countdata)
# Convert to matrix
countdata <- as.matrix(countdata)
head(countdata)
# Assign condition 
treat <- factor(c("CON","GFP","CON","GFP","CON","GFP"))

# Create a coldata frame and instantiate the DESeqDataSet
coldata <- data.frame(row.names=colnames(countdata), treat)

dds <- DESeqDataSetFromMatrix(countData = countdata,
                              colData = coldata,
                              design= ~ treat)

#filter low gene counts
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds

# Run the DESeq pipeline
dds <- DESeq(dds)
colnames(dds) <- colnames(countdata)
#contrast GFP over control
res <- results(dds, contrast=c('treat','GFP','CON')) 
dim(res)
res <- res[complete.cases(res),]  #remove any rows with NA
dim(res)
result <- res[order(res$padj), ]
head(result)
table(result$padj<0.05)

#print results with norm counts
resdata <- merge(as.data.frame(result), as.data.frame(counts(dds, normalized=TRUE)), by="row.names", sort=FALSE)
names(resdata)[1] <- "Gene"
head(resdata)
## Write results
write.csv(resdata, file="061620_DEG_n3_reactivation.csv")
```

```{r heatmaps}
#compare sample correlation via heatmap
library(pheatmap)
rld <- rlog(dds, blind=TRUE)
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)
pheatmap(rld_cor)

#compare sample distance vsd via heatmap
library(RColorBrewer)
vsd <- vst(dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

```

```{r tree}
# tree relationships
library(ape)
d <- cor(assay(rlog(dds)), method="spearman")
hc <- hclust(dist(1-d))
plot.phylo(as.phylo(hc), type="p", edge.col=4, edge.width=3, show.node.label=TRUE, no.margin=TRUE)

```

```{r pca}
# PCA style1
library("PoiClaClu")
plotPCA(vsd, intgroup = "treat")
```

```{r volcano}
#volcano plot
library(ggrepel)
results = as.data.frame(dplyr::mutate(as.data.frame(result), sig=ifelse(result$padj<0.05, "FDR<0.05", "Not Sig")), row.names=rownames(result))
head(results)

DEgenes_DESeq <- results[which(abs(results$log2FoldChange) > log2(1) & results$padj < 0.05),]

p <-  ggplot2::ggplot(results, ggplot2::aes(log2FoldChange, -log10(pvalue))) +
  ggplot2::geom_point(ggplot2::aes(col = sig)) +
  ggplot2::scale_color_manual(values = c("red", "black")) +
  ggplot2::ggtitle("reactivation")

p
p + ggrepel::geom_text_repel(data=results[1:20, ], ggplot2::aes(label=rownames(results[1:20, ])))

```

```{r plotgenecounts}
#plot counts
Apln = plotCounts(dds, gene="Apln", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene1 <- Apln + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Apln expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

Hdac1 = plotCounts(dds, gene="Hdac1", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene2 <- Hdac1 + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Hdac1 expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

Hspa1b = plotCounts(dds, gene="Hspa1b", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene3 <- Hspa1b + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Hspa1b expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

Hspb11 = plotCounts(dds, gene="Hspb11", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene4 <- Hspb11 + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Hspb11 expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

Hsp90aa1 = plotCounts(dds, gene="Hsp90aa1", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene5 <- Hsp90aa1 + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Hsp90aa1 expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

Hspa1a = plotCounts(dds, gene="Hspa1a", intgroup = "treat", returnData = TRUE) %>%
  ggplot() + aes(treat, count) + geom_boxplot(aes(fill=treat))  + theme_bw()

gene6 <- Hspa1a + geom_jitter(shape=16, position=position_jitter(0.2)) + labs(title="Hspa1a expression") + theme_classic() + theme(plot.title = element_text(hjust = 0.5))

library(grid)
library(ggpubr)
plot <- ggarrange(gene1, gene2, gene3, gene4, gene5, gene6, nrow=3, ncol=3, common.legend = TRUE)
plot
ggsave("gene_counts_n3.pdf")

```

```{r biomart}
#pathway analsyis convert to entrez gene id
library(biomaRt)
ensembl <- useMart("ensembl", dataset="mmusculus_gene_ensembl")
bm <- getBM(attributes=c("ensembl_gene_id","entrezgene_id",
                         "mgi_symbol"),
            filters="mgi_symbol",
            values=resdata$Gene,
            mart=ensembl)
head(bm)
#rename mgi_symbol to Gene to merge with resdata
names(bm)[3] <- "Gene"
#merge bm output with resdata
merged_data <- merge(resdata, bm, by="Gene", all.x=T, sort=F)
head(merged_data)

```

```{r ranks}
#filter by significance
#sig_genes <- merged_data[abs(merged_data$log2FoldChange)>=1 & (merged_data$padj < 0.05),]

resSig <- subset(merged_data, padj < 0.1)
head(resSig)
dim(resSig)

#make ranked tbl_df with entrezgene_id stat metric, remove na
#summarize the mean stat across different gene ids with nas
rank_stat <- resSig %>% 
  dplyr::select(entrezgene_id, stat) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(entrezgene_id)%>% 
  summarize(stat=mean(stat))
head(rank_stat)
dim (rank_stat)
#make ranked tbl_df with lfc metric
rank_lfc <- resSig %>% 
  dplyr::select(entrezgene_id,log2FoldChange) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  summarize(log2FoldChange=mean(log2FoldChange))
head(rank_lfc)

#check the results
ResTidy <- resSig %>%
  as_tibble() %>%
  arrange(desc(padj))
#view in table
ResTidy %>% 
  dplyr::select(Gene, entrezgene_id, padj, log2FoldChange) %>% 
  arrange(padj) %>% 
  DT::datatable()

stat <- deframe(rank_stat)
head(stat, 20)
barplot(sort(stat, decreasing = T))

lfc <- deframe(rank_lfc)
head(lfc, 20)
barplot(sort(lfc, decreasing = T))
```

```{r fgsea}
library(fgsea)
library(limma)
#GSEA hallmarks gene set
Mm.c2 <- readRDS(url("http://bioinf.wehi.edu.au/MSigDB/v7.1/Mm.c2.all.v7.1.entrez.rds"))
Mm.c5 <- readRDS(url("http://bioinf.wehi.edu.au/MSigDB/v7.1/Mm.c5.all.v7.1.entrez.rds"))
pathways.c2 <- Mm.c2
pathways.c5 <- Mm.c5
#stat metric
fgseastat.c2 <- fgseaMultilevel(pathways.c2, stat, minSize=10, maxSize = 500)
fgseastat.c5 <- fgseaMultilevel(pathways.c5, stat, minSize=10, maxSize = 500)
#lfc metric
fgsealfc.c2 <- fgseaMultilevel(pathways.c2, lfc, minSize=10, maxSize = 500)
fgsealfc.c5 <- fgseaMultilevel(pathways.c5, lfc, minSize=10, maxSize = 500)
#tidy the results

p <-  ggplot(fgsealfc.c2, aes(x=NES, y=pathway, size=padj,color=-1*log10(padj))) + geom_point()+scale_size_area(max_size = 5)+scale_colour_gradient(low="green",high="red")
ggplot(fgsealfc.c2, aes(x="Name of sample", y=pathway, size=padj,color=NES)) + geom_point()+scale_size_area(max_size = 5)+scale_colour_gradient(low="green",high="red")
fgseaResTidystat.c2 <- fgseastat.c2 %>%
  as_tibble() %>%
  arrange(desc(NES))

fgseaResTidylfc.c2 <- fgsealfc.c2 %>%
  as_tibble() %>%
  arrange(desc(NES))

fgseaResTidystat.c5 <- fgseastat.c5 %>%
  as_tibble() %>%
  arrange(desc(NES))

fgseaResTidylfc.c5 <- fgsealfc.c5 %>%
  as_tibble() %>%
  arrange(desc(NES))

# Show in tables:
fgseaResTidystat.c2 %>% 
  dplyr::select(pathway, padj, NES) %>% 
  arrange(padj) %>% 
  DT::datatable()

fgseaResTidylfc.c2 %>% 
  dplyr::select(pathway, padj, NES) %>% 
  arrange(padj) %>% 
  DT::datatable()

fgseaResTidystat.c5 %>% 
  dplyr::select(pathway, padj, NES) %>% 
  arrange(padj) %>% 
  DT::datatable()

fgseaResTidylfc.c5 %>% 
  dplyr::select(pathway, padj, NES) %>% 
  arrange(padj) %>% 
  DT::datatable()


plotEnrichment(pathways.c5[["GO_integral_component_of_membrane"]], stat)
```

```{r clusterprofiler}
# CLUSTERPROFILER
#log2 fold change of padj < 0.1 sig genes
ResTidy
cp_genes <- resSig %>% 
  dplyr::select(entrezgene_id, log2FoldChange) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(entrezgene_id)
head(cp_genes)

df<- cp_genes$log2FoldChange

# name the vector
names(df) <- cp_genes$entrezgene_id

# omit any NA values 
gene_list<-na.omit(df)

# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)
organism = "org.Mm.eg.db"
library(organism, character.only = TRUE)
library(clusterProfiler)

library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]


# library(DOSE)
# gse <- gseGO(geneList=gene_list, 
#              ont ="ALL", 
#              keyType = "ENTREZID", 
#              minGSSize = 10, 
#              maxGSSize = 800,
#              eps = 1e-10,
#              pvalueCutoff = 0.05,
#              verbose = TRUE, 
#              OrgDb = "org.Mm.eg.db", 
#              pAdjustMethod = "BH")
# 
# require(DOSE)
# dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)

```

